Men and past smokers from Ontario are potentially more physically active than women and non-smokers

### 2024-12-08 paper/paper.qmd
Final updates to paper
1. Shortened subtitle.
2. Corrected grammar and spelling.
3. Added more weaknesses of paper.
4. Added to telegraphing paragraph to include appendix B.
5. Added justification to 'Variable 1: Age' section as to why age would be included in model exploration.
6. Changed 'Histogram of', 'Boxplot of' 'Barplot of' to other terminology.

---
title: "Men and past smokers from ontario are potentially more physically active than women and non-smokers"
subtitle: "Women are engaging in lesser physical activities than men. Furthermore, individuals who smoked before are more likely to engage in physical activity. Consumption of stimulant-related drugs could have also increased as well, resulting in increased overall physical activities in Ontario."
author: 
  - Chris Yong Hong Sen
thanks: "Code and data are available at: [https://github.com/Monoji77/Alcohol_Use_Ontario](https://github.com/Monoji77/Alcohol_Use_Ontario)."
date: today
date-format: long
abstract: "This study explores factors influencing physical activity levels in Ontario, focusing on demographic and substance abuse. Our findings reveal that women engage in less physical activity than men, while individuals who have smoked in the past and those consuming stimulant-related drugs are more likely to be physically active. Additionally, age, highest educational attainment, and health region also play significant roles in determining physical activity levels. These insights highlight the complex interplay of gender, lifestyle behaviors, and sociodemographic factors, providing valuable information for designing targeted public health interventions to promote physical activity."
format: 
  pdf:
    fig-pos: H
number-sections: true
bibliography: references.bib
---

@@

Recent studies suggest that certain demographic groups, such as older adults and individuals with lower educational levels, are more likely to experience barriers to physical activity, while alcohol consumption and smoking are associated with lower physical activity levels [@citationBaumanPhysicalActivity]. However, limited research has explored how these factors specifically interact in Ontario, a province with diverse populations and health regions. As the burden of chronic disease grows, addressing these disparities becomes even more urgent [@citationWarburtonPhysicalActivity] .

@@

The remainder of this paper is structured as follows. @sec-data would provide a sense of the chosen Computing in the Humanities and Social Sciences dataset as well as data exploration of interested variables. @sec-Model goes into the details of the linear regression model fitted on the predictors of interest variables, for which it's results would be showcased in @sec-results. Finally @sec-discussion would interpret the results of the fitted linear regression model, which includes limitations of the paper. At the end of the paper in [Appendix -@sec-diagnostic_checks], this contains diagnostic checks for all models considered.

Overall, this paper was generated using R programming language [@citeR]. Reading of raw data and data manipulation was performed using `Tidyverse` package [@citeTidyverse] and parquet file handling was done using functions in `arrow` package [@citeArrow]. Generation of maps were done using functions in `tmap` package [@citationTmap]. Any other visual plots were using `ggplot`-related functions in Tidyverse [@citeTidyverse] and tables were generated using `kable` function in `knitr` package [@citeKnitR].

@@

The dataset chosen is obtained from Computing in the Humanities and Social Sciences/CHASS at the university of Toronto [@SDA_CHASS], specifically the Canadian Community Health Survey (CCHS) annual component 2017-2018 which was collected by @StatisticsOfCanada. The goal of the CCHS survey is to collect comprehensive and reliable health-related data on the Canadian population to support health surveillance, research, and program evaluation. It aims to provide timely, accessible, and flexible data for monitoring health trends, studying small populations and rare characteristics, and addressing emerging health issues, ultimately helping improve the health and well-being of Canadians.

Other datasets explored include sources from World Health Organisation and ParticipACTION. World Health Organisation provides aggregated data informing certain indicators of interest such as obesity levels and alcohol consumption. This can be found in the data section for Canada[@OtherSource1]. Also, ParticipACTION provides key statistics regarding proportion of adult population in canada meeting the national guidelines of 150 minutes of moderate to vigorous physical activites, as well as how sedentary canadians are [@OtherSource2]. However, the availability of raw individual level data is absent from these sources. Although the aggregated data gives us a sensing about general Canadian health, we are not able to focus our attention to the Ontario population. The dataset chosen from Canadian Community Health Survey (CCHS) annual component is ideal as it contains individual level data, where we are able to obtain the province the respondents are from.

@@

The dataset provides 113,290 individual response from Canadians all throughout the country. Surveys were conducted to collect health-related data from respondents. The relevant data collected includes...

@@

This dataset is highly suitable in determining both quantitative factors such as  c qualitative factors

@@

According to the questionnaire, respondents were asked in a phone survey to indicate whether they had engaged in any strenuous activity that lasted at least 10 minutes and made them sweat and breathe a little harder in the past 7 days. Afterwards, they were asked to indicate the total time spent on those activities. To address the concern of recall bias where an individual may not recall the specific timing in engaging with physical activity, surveyors sectioned physical activity into commuting, sports, work and volunteering. Then within each section, surveyors asked which day (Monday to Sunday) in the past week had they engaged in such activities that made them sweat and breathe a little harder. By doing so, the activities individuals did during the past week would be anchored on the days of the week. Following this would be for the respondents to indicate the total number of hours they engaged in each section of physical activities for the past 7 days.

@@

For socio-demographic factors possibly affecting time spent on physical activites, the surveyors asked the respondent for their age (DHH_AGE) from 0 years old to 121 years old), sex (DHH_SEX) being either male or female, personal income obtained from tax forms submitted to the Canada Revenue Agency upon consent from respondent, geographical health region (ADM_Q037) that was provided by respondents when they agreed to the phone survey, and highest educational attainment (EHG2_04) ranging from less than high school diploma to master's degree and above.

For substance abuse, respondents were asked how frequent they consumed alcoholic drinks in the past 12 months (ALC_Q015), ranging from less than once per month to everyday. Surveyors also asked respondents for usage of drugs including one-time marijuana or hashish in the past 12 months (DRG_Q005 to DRG_Q075), and whether they have smoked hundred cigarettes in the past 12 months (SMK_Q020). The last 2 questions allowed respondents to indicate a simple yes or no flag to express their answer.

For health factors, questions were asked to determine an individual's body mass index ranging from underweight to obese and perceived life/work stress ranging from not at all stressful to extremely stressful. Additionally, surveyors asked respondents if they have mood disorders (E.g. depression, bipolarism, mania), for which respondents replied with a simple yes or no.

Overall, each entry obtained by respondents assigns quantitative values to each factor and the the response for which we can explore subsequently using linear regression to determine which factors have greater influence in the time spent on physical activities amongst active individuals.

@@

Since we are interested in individuals that are already engaging in strenuous activities, we can filter $time\_spent\_vigorous\_exercise\_7d$ to exclude 0. After removing such entries, we obtain the following summary statistics as shown in @tbl-summarised_statistics_time_spent_vigorous_exercise_7d. The median time spent on physical activities the past 7 days would be 300 minutes, which averages out to 43 minutes per day.

@@

Interestingly, the maximum time spent on physical activities the past 7 days was 9902 minutes, and this averages to 1415 minutes per day, which is equivalent to 23.5 hours in a day. This is very unlikely as this suggests that an individual is constantly engaging in physical strenuous activity without sleeping. Logically, this suggests the possibility of outlier values. By observing @fig-boxplot_time_spent_vigorous_exercise_7d, there are many outlier observations for this variable. Roughly 10% of the observed data are considered outliers as they fall outside 1.5 times the interquartile range from the 25% and 75% quartile values. We would be performing log transformation on our response variable. The motivation for this is detailed in [Appendix -@sec-transform-response].

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boxplot_time_spent_vigorous_exercise_7d
#| fig-cap: Boxplot of time spent on physical activities in the past 7 days

@@

By observing the histogram generated in @fig-hist-time_spent_vigorous_exercise_7d, we obtain a severely right skewed response histogram where most observations lie between 0 to 500 minutes. For the building of linear models in subsequent sections of the paper, this would not be ideal as linear regression assumes a normal distribution of the response variable, suggesting that a transformation would be required to fit a linear regression model.

```{r response variable exploration}
#| echo: false
#| warning: false
#| message: false
#| label: fig-hist-time_spent_vigorous_exercise_7d
#| fig-cap: Histogram of time spent on physical activities the past 7 days

hist(data_cleaned_1$time_spent_vigorous_exercise_7d,
     main='',
     xlab = '',
     ylab = 'no. of individuals',
     freq=TRUE, cex.main=1, las = 1
    )
```
## Interested explanatory variables {#sec-interested_variables}

There are a total of 13 predictors of interest that is seemingly related to the time spent on physical activities by an individual in Ontario.

### Variable 1: Age
This is a variable of interest because age is seemingly correlated to The age groups of respondents were obtained and they exist in bins that spans ages 18 years to 74 years old.

@@

From @fig-barplot-age, we can immediately see that most observations fall within age groups 30 to 34 years old (bin 6), and lesser observations fall within age groups 70-74 years old (bin 14).
```{r age predictor barplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot-age
#| fig-cap: Histogram of time spent on physical activites the past 7 days

@@

From @tbl-summarised_statistics_age and @fig-boxplot-age, we are able to see that the the median age bin belongs to age group 40-44 years old. We are able to see that there are no outlier values from @fig-boxplot-age, which means that the survey has obtained an acceptable number of respondents across all age bins. 
```{r age predictor summarised statistics}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summarised_statistics_age
#| tbl-cap: Summarised statistics of respondent ages

@@

```{r age predictor boxplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boxplot-age
#| fig-cap: Boxplot of respondent ages

@@

Sex is a possible influential predictor for total time spent on physical activity. Historically, men are associated with exercise and involved in more labour intensive roles such as construction which involves heavy lifting. According to @womenExerciseWHO, women are less active than men by 5% since 2000. However, with the advent of female only establishments such as female gyms, and 32% increase in memberships at fitness and health clubs from 2010 to 2019, it would be interesting to see if the sex of an individual still affects the time spent on physical activites among physically active individuals.

According to @fig-barplot-sex, there is roughly an equal proportion of sexes from our dataset, with slightly more males (6082 respondents) than females (5859 respondents). Since this was a health survey conducted over phone call, there is an equal chance of a respondent to be male or female, which is roughly reflected ing @fig-barplot-sex.
```{r sex predictor barplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot-sex
#| fig-cap: Barplot of sex
data_cleaned_1 |>
  mutate(sex = factor(sex, levels = c(0,1), labels =c('Female','Male'))) |>
  ggplot()  +
  geom_bar(aes(x=sex, fill=sex), show.legend=F) +
  scale_fill_manual(values = c('Female' = 'indianred4',
                               'Male' = 'skyblue1')) +
  theme_minimal() +
  xlab('Individual\'s sex') +
  ylab('Number of individuals') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

@@

In 2023, @EducationAndActiveLifestyle has done a study to show that highest educational attainment may be a leading factor for physical activity. Therefore, it would be worth exploring the effects of highest levels of education with the time spent on physical activities among the Ontario residents.

According to @tbl-proportion_highest_educational_attainment, respondents who did not graduate secondary school account for only 5% of the population, and respondents who have at least a bachelor's degree account for 72% of the population. According to @fig-barplot_highest_educational_attainment, there seems to be class imbalance for this variable.

However, as seen from post-secondary institution enrollment data from @statisticaPostSecEnrollment, there has been an increase in total number of Canadians enrolled in post secondary education from 2000 to 2020. Accounting for population changes from 2000 [@statCanadaPopulation2000] to 2022 [@statCanadaPopulation2022], there is still a real increase in the number of real enrollments into post secondary education. Therefore, it is very possible that this is representative of the Ontario's population where majority of the population has at least post secondary education.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_highest_educational_attainment
#| fig-cap: Barplot of highest_educational_attainment
data_cleaned_1 |>
  mutate(highest_educational_attainment = factor(highest_educational_attainment, levels = c(1,2,3), labels =c('less than sec sch grad','sec sch grad', 'university degree and above'))) |>
  ggplot()  +
  geom_bar(aes(x=highest_educational_attainment, fill=highest_educational_attainment), show.legend=F) +
  scale_fill_manual(values = c('less than sec sch grad' = 'orange1',
                               'sec sch grad' = 'indianred4',
                               'university degree and above' = 'skyblue1')) +
  theme_minimal() +
  xlab('Highest educational attainment') +
  ylab('Number of individuals') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

@@

Health region is a categorical variable which tells us the geographical location an individual belongs to defined by Provincial Ministries of Health (as detailed by documentation provided by CHASS). It could be possible that certain health regions have more funding that are projected towards increasing awareness towards having an active lifestyle. Therefore, it is worth exploring the relationship between health region and time spent on physical activities.

There are a total of 34 different health regions in the Ontario province. They are encoded using 5 digits. @fig-map_health_region shows us the various health regions in Ontario. The blue regions shows us health regions for which lesser respondents belong to while red regions show us health regions for which more respondents belong to.

@@

By using a GIS focused library in R called tmap [@citationTmap], we are able to generate the map shown in @fig-map_health_region.

@@

Overall, @fig-barplot_health_region shows us the actual number of observations belonging to the top five and bottom 5 occurring health regions. We can see that most respondents come from the City of Ontario Health Unit (Health_Region = 35995) and lesser respondents belong to Porcupine Health Unit (Health_Region = 35956).

```{r Health region map}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_health_region
#| fig-cap: Barplot of top and bottom five occuring health regions

@@

In the dataset, personal income (INCDGPER) is a categorical variable that places an individuals annual income into a range of income values. Below summarises the encoded number and the corresponding meaning of each encoded number. All income values are measured per year.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_personal_income
#| fig-cap: Barplot of personal income

@@

From @fig-barplot_frequency_alcohol_consumption, we are able to see that most respondents from this dataset drink about 2-3 times per week (2973 respondents) as seen from the red bar. Also, we are seeing fewer respondents drinking about 4-6 times per week (887 respondents) as seen from the blue bar.
```{r Alcohol frequency bar graph}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_frequency_alcohol_consumption
#| fig-cap: Barplot of frequency of alcohol consumption
max_alc <- data_cleaned_1 |> 
  count(num_alc_drank_12m) |> 
  top_n(1, n) |>
  pull(num_alc_drank_12m)

@@

This is a binary variable which indicates whether an individual has used drugs in the past 12 months prior to the interview. Drugs include marijuana, hashish, cocaine, amphetamines, speed, methamphetamine, crystal meth, ecstasy, hallucinogens, and sniffing glue, gasoline or other solvents. If they have attempted any of these drugs, this variable would be flagged as '1' and '0' otherwise.

It would be interesting to see if drug consumption has an effect on physical activeness. @PhysicalActivityAndSubstanceAbuse exerts that individuals who are more active tend to have lower substance abuse, suggesting a correlation between these two. Having this variable would be interesting to explore the relationship of an individuals drug usage and their time spent on physical activities.

@fig-barplot_illicit_drug_use shows us the number of respondents who have (or haven't) consumed drugs 12 months prior to their interviews. We can see that majority of respondents have not consumed drugs in the given timeframe. However, from @tbl-illicit_drug_use, we can see that there are still a number of respondents who had consumed drugs in the past 12 months, particularly a fifth of the sampling population. According to @SubstanceUseAndAddiction, they estimate that about 21% of canadians, or 6 million people, will meet the criteria for substance use in their lifetime, and this is inclusive of alcohol and that there is a growing opiod crisis in Ontario. Therefore the obtained sample is somewhat representative of usage of drugs among the Ontario population. 
```{r Illicit drug use summary table}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-illicit_drug_use
#| tbl-cap: Summary of illicit drug use the past 12 months

@@

```{r Illicit drug use barplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_illicit_drug_use
#| fig-cap: Barplot of illicit drug use the past 12 months
data_cleaned_1 |>
  mutate(illicit_drug_use = factor(illicit_drug_use, levels = c(0,1), labels =c('Non-drug user','Drug user'))) |>
  ggplot()  +
  geom_bar(aes(x=illicit_drug_use, fill=illicit_drug_use), show.legend=F) +
  scale_fill_manual(values = c('Non-drug user' = 'indianred4',
                               'Drug user' = 'skyblue1')) +
  theme_minimal() +
  xlab('') +
  ylab('Number of respondents') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

@@

Respondents were asked whether they had smoked more than 100 cigarettes in their lifetimes. This variable is therefore binary, with '1' representing respondents that meet the critera and '0' otherwise.

It is well known that excessive smoking has negative health impacts on individuals. A cross sectional study conducted by @SmokingAndPhysicalActiveness found that smokers are less likely to be physically active compared to non-smokers. Hence, it would be interesting to discover if there is an effect of smoking more than 100 cigarettes on the time spent on physical activites.

@fig-barplot-cigarettes shows us that within the sample, the visual difference between the number of respondents who smoked more than 100 cigarettes (blue column) and respondents who smoked lesser than 100 cigarettes are not that big. Specifically, 57% of respondents have smoked less than 100 cigarettes while the remaining 43% of respondents smoked more than 100 cigarettes. As the classes in this variable is almost split half,

@@

Initially, we fitted a full model containing every variable of interest. Since health region is a categorical variable with no implied order, To handle health region, we will be creating dummy variables for each health region, specifically $I*(HealthRegion_k=i)$ for $i \in {health\_region}$, and the reference chosen for health region would be the health_region = 35926 (The District of Algoma HU). We define $Income_k$ to be the the personal income of individual k in our dataset.

@@

Overall, the individual t-test for our predictors were found to be below our set threshold of 95% alpha. The confidence interval for each predictor is listed in @tbl-confidence_interval_lmmodel shown below.

>>>>

---
title: "Men and past smokers from Ontario are potentially more physically active than women and non-smokers"
subtitle: "Physical activity is discovered to be strongly associated with gender, smoking habits, and drug consumption."
author: 
  - Chris Yong Hong Sen
thanks: "Code and data are available at: [https://github.com/Monoji77/Alcohol_Use_Ontario](https://github.com/Monoji77/Alcohol_Use_Ontario)."
date: today
date-format: long
abstract: "This study explores factors influencing physical activity levels in Ontario, focusing on demographics and substance abuse. Our findings reveal that women engage in less physical activity than men, while individuals who have smoked in the past and those consuming stimulant-related drugs are more likely to be physically active. Additionally, age, highest educational attainment, and health region also play significant roles in determining physical activity levels. These insights highlight the complex interplay of gender, lifestyle behaviors, and sociodemographic factors, providing valuable information for designing targeted public health interventions to promote physical activity."
format: 
  pdf:
    fig-pos: H
number-sections: true
bibliography: references.bib
---

@@

Recent studies suggest that certain demographic groups, such as older adults and individuals with lower educational levels, are more likely to experience barriers to physical activity. At the same time, alcohol consumption and smoking are associated with lower physical activity levels [@citationBaumanPhysicalActivity]. However, limited research has explored how these factors specifically interact in Ontario, a province with diverse populations and health regions. As the burden of chronic disease grows, addressing these disparities becomes even more urgent [@citationWarburtonPhysicalActivity].

@@

The remainder of this paper is structured as follows. @sec-data would provide a sense of the chosen Computing in the Humanities and Social Sciences dataset as well as data exploration of interested variables. @sec-Model goes into the details of the linear regression model fitted on the predictors of interest variables, for which its results would be showcased in @sec-results. @sec-discussion would then interpret the results of the fitted linear regression model, which includes the limitations of the paper. In [Appendix -@sec-diagnostic_checks], it contains diagnostic checks for all models considered. Finally, [Appendix -@sec-survey_methodology] goes through the background and surveying methodology of the CCHS dataset.

Overall, this paper was generated using R programming language [@citeR]. Reading of raw data and data manipulation was performed using `Tidyverse` package [@citeTidyverse] and parquet file handling was done using functions in `arrow` package [@citeArrow]. Generation of maps was done using functions in `tmap` package [@citationTmap]. Any other visual plots were using `ggplot`-related functions in Tidyverse [@citeTidyverse] and tables were generated using `kable` function in `knitr` package [@citeKnitR].

@@

The dataset chosen is obtained from Computing in the Humanities and Social Sciences/CHASS at the University of Toronto [@SDA_CHASS], specifically the Canadian Community Health Survey (CCHS) annual component 2017-2018 which was collected by @StatisticsOfCanada. The goal of the CCHS survey is to collect comprehensive and reliable health-related data on the Canadian population to support health surveillance, research, and program evaluation. It aims to provide timely, accessible, and flexible data for monitoring health trends, studying small populations and rare characteristics, and addressing emerging health issues, ultimately helping improve the health and well-being of Canadians.

Other datasets explored include sources from the World Health Organisation and ParticipACTION. World Health Organisation provides aggregated data informing certain indicators of interest such as obesity levels and alcohol consumption. This can be found in the data section for Canada[@OtherSource1]. Also, ParticipACTION provides key statistics regarding the proportion of the adult population in Canada meeting the national guidelines of 150 minutes of moderate to vigorous physical activities, as well as how sedentary Canadians are [@OtherSource2]. However, the availability of raw individual-level data is absent from these sources. Although the aggregated data gives us a sense of general Canadian health, we are not able to focus our attention on the Ontario population. The dataset chosen from the Canadian Community Health Survey (CCHS) annual component is ideal as it contains individual-level data, where we are able to obtain the province the respondents are from in order to obtain our desired target population.

@@

The dataset provides 113,290 individual responses from Canadians throughout the country. Surveys were conducted to collect health-related data from respondents. The relevant data collected includes...

@@

This dataset is highly suitable for determining both quantitative factors such as  c qualitative factors.

@@

According to the questionnaire, respondents were asked in a phone survey to indicate whether they had engaged in any strenuous activity that lasted at least 10 minutes and made them sweat and breathe a little harder in the past 7 days. Afterward, they were asked to indicate the total time spent on those activities. To address the concern of recall bias where an individual may not recall the specific timing in engaging with physical activity, surveyors sectioned physical activity into commuting, sports, work, and volunteering. Then within each section, surveyors asked which day (Monday to Sunday) in the past week had they engaged in such activities that made them sweat and breathe a little harder. By doing so, the activities individuals did during the past week would be anchored on the days of the week. Following this would be for the respondents to indicate the total number of hours they engaged in each section of physical activities for the past 7 days.

@@

For socio-demographic factors possibly affecting time spent on physical activities, the surveyors asked the respondent for their age (DHH_AGE) from 0 years old to 121 years old), sex (DHH_SEX) being either male or female, personal income obtained from tax forms submitted to the Canada Revenue Agency upon consent from the respondent, geographical health region (ADM_Q037) that was provided by respondents when they agreed to the phone survey, and highest educational attainment (EHG2_04) ranging from less than high school diploma to master's degree and above.

For substance abuse, respondents were asked how frequently they consumed alcoholic drinks in the past 12 months (ALC_Q015), ranging from less than once per month to every day. Surveyors also asked respondents about the usage of drugs including one-time marijuana or hashish in the past 12 months (DRG_Q005 to DRG_Q075), and whether they have smoked hundred cigarettes in the past 12 months (SMK_Q020). The last 2 questions allowed respondents to indicate a simple yes or no flag to express their answer.

For health factors, questions were asked to determine an individual's body mass index ranging from underweight to obese, and perceived life/work stress ranging from not at all stressful to extremely stressful. Additionally, surveyors asked respondents if they have mood disorders (E.g. depression, bipolar, mania), to which respondents replied with a simple yes or no.

Overall, each entry obtained by respondents assigns quantitative values to each factor and the the response for which we can explore subsequently using linear regression to determine which factors have greater influence on the time spent on physical activities amongst active individuals.

@@

Since we are interested in individuals who are already engaging in strenuous activities, we can filter $time\_spent\_vigorous\_exercise\_7d$ to exclude 0. After removing such entries, we obtain the following summary statistics as shown in @tbl-summarised_statistics_time_spent_vigorous_exercise_7d. The median time spent on physical activities the past 7 days would be 300 minutes, which averages out to 43 minutes per day.

@@

Interestingly, the maximum time spent on physical activities in the past 7 days was 9902 minutes, and this averages to 1415 minutes per day, which is equivalent to 23.5 hours in a day. This is very unlikely as this suggests that an individual is constantly engaging in physical strenuous activity without sleeping. Logically, this suggests the possibility of outlier values. By observing @fig-boxplot_time_spent_vigorous_exercise_7d, there are many outlier observations for this variable. Roughly 10% of the observed data are considered outliers as they fall outside 1.5 times the interquartile range from the 25% and 75% quartile values. We would be performing log transformation on our response variable. The motivation for this is detailed in [Appendix -@sec-transform-response].

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boxplot_time_spent_vigorous_exercise_7d
#| fig-cap: Summary statistic of time spent on physical activities in the past 7 days

@@

By observing @fig-hist-time_spent_vigorous_exercise_7d, we obtain a severely right-skewed response where most observations lie between 0 to 500 minutes. For the building of linear models in subsequent sections of the paper, this would not be ideal as linear regression assumes a normal distribution of the response variable, suggesting that a transformation would be required to fit a linear regression model.

```{r response variable exploration}
#| echo: false
#| warning: false
#| message: false
#| label: fig-hist-time_spent_vigorous_exercise_7d
#| fig-cap: Time spent on physical activities the past 7 days

hist(data_cleaned_1$time_spent_vigorous_exercise_7d,
     main='',
     xlab = 'no. of minutes',
     ylab = 'no. of individuals',
     freq=TRUE, cex.main=1, las = 1
    )
```
## Interested explanatory variables {#sec-interested_variables}

There are a total of 13 predictors of interest that are seemingly related to the time spent on physical activities by an individual in Ontario.

### Variable 1: Age
This is a variable of interest because age is seemingly correlated to time spent on physical activities. It is well known that the younger schooling population has more time and physical capacity to exercise than a working adult. In the International Journal of Behavioral Nutrition and Physical Activity, @citationUofExerciseAndAge discovered that younger individuals were exercising more than their older counterparts.

The age groups of respondents were obtained and they exist in bins that span ages 18 years to 74 years old.

@@

From @fig-barplot-age, we can immediately see that most observations fall within age groups 30 to 34 years old (bin 6), and lesser observations fall within age groups 70-74 years old (bin 14).
```{r age predictor barplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot-age
#| fig-cap: Time spent on physical activities the past 7 days

@@

From @tbl-summarised_statistics_age and @fig-boxplot-age, we are able to see that the median age bin belongs to the age group 40-44 years old. We are able to see that there are no outlier values from @fig-boxplot-age, which means that the survey has obtained an acceptable number of respondents across all age bins. 
```{r age predictor summarised statistics}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-summarised_statistics_age
#| tbl-cap: Summarised statistics of respondent ages

@@

```{r age predictor boxplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-boxplot-age
#| fig-cap: Summary statistics of respondent ages

@@

Sex is a possible influential predictor for total time spent on physical activity. Historically, men are associated with exercise and involved in more labour-intensive roles such as construction which involves heavy lifting. According to @womenExerciseWHO, women have been less active than men by 5% since 2000. However, with the advent of female-only establishments such as female gyms, and a 32% increase in memberships at fitness and health clubs from 2010 to 2019, it would be interesting to see if the sex of an individual still affects the time spent on physical activities among physically active individuals.

According to @fig-barplot-sex, there is roughly an equal proportion of sexes from our dataset, with slightly more males (6082 respondents) than females (5859 respondents). Since this was a health survey conducted over a phone call, there is an equal chance of a respondent being male or female, which is roughly reflected in @fig-barplot-sex.
```{r sex predictor barplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot-sex
#| fig-cap: Frequency of sex variable
data_cleaned_1 |>
  mutate(sex = factor(sex, levels = c(0,1), labels =c('Female','Male'))) |>
  ggplot()  +
  geom_bar(aes(x=sex, fill=sex), show.legend=F) +
  scale_fill_manual(values = c('Female' = 'indianred4',
                               'Male' = 'skyblue1')) +
  theme_minimal() +
  xlab('Individual\'s sex') +
  ylab('Number of individuals') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

@@

In 2023, @EducationAndActiveLifestyle did a study to show that the highest educational attainment may be a leading factor for physical activity. Therefore, it would be worth exploring the effects of the highest levels of education on the time spent on physical activities among Ontario residents.

According to @tbl-proportion_highest_educational_attainment, respondents who did not graduate secondary school account for only 5% of the population, and respondents who have at least a bachelor's degree account for 72% of the population. According to @fig-barplot_highest_educational_attainment, there seems to be a class imbalance for this variable.

However, as seen from post-secondary institution enrollment data from @statisticaPostSecEnrollment, there has been an increase in the total number of Canadians enrolled in post-secondary education from 2000 to 2020. Accounting for population changes from 2000 [@statCanadaPopulation2000] to 2022 [@statCanadaPopulation2022], there is still a real increase in the number of real enrollments in post-secondary education. Therefore, it is very possible that this is representative of Ontario's population where the majority of the population has at least post-secondary education.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_highest_educational_attainment
#| fig-cap: Frequency of highest_educational_attainment variable
data_cleaned_1 |>
  mutate(highest_educational_attainment = factor(highest_educational_attainment, levels = c(1,2,3), labels =c('less than sec sch grad','sec sch grad', 'university degree and above'))) |>
  ggplot()  +
  geom_bar(aes(x=highest_educational_attainment, fill=highest_educational_attainment), show.legend=F) +
  scale_fill_manual(values = c('less than sec sch grad' = 'orange1',
                               'sec sch grad' = 'indianred4',
                               'university degree and above' = 'skyblue1')) +
  theme_minimal() +
  xlab('Highest educational attainment') +
  ylab('Number of individuals') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())

@@

Health region is a categorical variable that tells us the geographical location an individual belongs to defined by Provincial Ministries of Health (as detailed by documentation provided by CHASS). It could be possible that certain health regions have more funding that are projected towards increasing awareness towards having an active lifestyle. Therefore, it is worth exploring the relationship between health region and time spent on physical activities.

There are a total of 34 different health regions in the Ontario province. They are encoded using 5 digits. @fig-map_health_region shows us the various health regions in Ontario. The blue regions show us health regions to which fewer respondents belong while the red regions show us health regions to which more respondents belong.

@@

By using a GIS-focused library in R called tmap [@citationTmap], we are able to generate the map shown in @fig-map_health_region.

@@

Overall, @fig-barplot_health_region shows us the actual number of observations belonging to the top five and bottom 5 occurring health regions. We can see that most respondents come from the City of Ontario Health Unit (Health_Region = 35995) and fewer respondents belong to the Porcupine Health Unit (Health_Region = 35956).

```{r Health region map}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_health_region
#| fig-cap: Frequency of top and bottom five occuring health regions

@@

In the dataset, personal income (INCDGPER) is a categorical variable that places an individual's annual income into a range of income values. Below summarises the encoded number and the corresponding meaning of each encoded number. All income values are measured per year.

@@

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_personal_income
#| fig-cap: Frequency of personal income

@@

From @fig-barplot_frequency_alcohol_consumption, we can see that most respondents from this dataset drink about 2-3 times per week (2973 respondents) as seen from the red bar. Also, we are seeing fewer respondents drinking about 4-6 times per week (887 respondents) as seen from the blue bar.
```{r Alcohol frequency bar graph}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_frequency_alcohol_consumption
#| fig-cap: Frequency of alcohol consumption
max_alc <- data_cleaned_1 |> 
  count(num_alc_drank_12m) |> 
  top_n(1, n) |>
  pull(num_alc_drank_12m)

@@

This is a binary variable that indicates whether an individual has used drugs in the past 12 months prior to the interview. Drugs include marijuana, hashish, cocaine, amphetamines, speed, methamphetamine, crystal meth, ecstasy, hallucinogens, and sniffing glue, gasoline or other solvents. If they have attempted any of these drugs, this variable would be flagged as '1' and '0' otherwise.

It would be interesting to see if drug consumption affects physical activity. @PhysicalActivityAndSubstanceAbuse exerts that individuals who are more active tend to have lower substance abuse, suggesting a correlation between these two. Having this variable would be interesting to explore the relationship between an individual's drug usage and their time spent on physical activities.

@fig-barplot_illicit_drug_use shows us the number of respondents who have (or haven't) consumed drugs 12 months before their interviews. We can see that the majority of respondents have not consumed drugs in the given timeframe. However, from @tbl-illicit_drug_use, we can see that there are still several respondents who had consumed drugs in the past 12 months, particularly a fifth of the sampling population. According to @SubstanceUseAndAddiction, they estimate that about 21% of Canadians, or 6 million people, will meet the criteria for substance use in their lifetime, and this is inclusive of alcohol and there is a growing opioid crisis in Ontario. Therefore the obtained sample is somewhat representative of the usage of drugs among the Ontario population. 
```{r Illicit drug use summary table}
#| echo: false
#| warning: false
#| message: false
#| label: tbl-illicit_drug_use
#| tbl-cap: Summary of illicit drug use the past 12 months

@@

```{r Illicit drug use barplot}
#| echo: false
#| warning: false
#| message: false
#| label: fig-barplot_illicit_drug_use
#| fig-cap: Illicit drug use the past 12 months
data_cleaned_1 |>
  mutate(illicit_drug_use = factor(illicit_drug_use, levels = c(0,1), labels =c('Non-drug user','Drug user'))) |>
  ggplot()  +
  geom_bar(aes(x=illicit_drug_use, fill=illicit_drug_use), show.legend=F) +
  scale_fill_manual(values = c('Non-drug user' = 'indianred4',
                               'Drug user' = 'skyblue1')) +
  theme_minimal() +
  xlab('') +
  ylab('Number of respondents') +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

@@

Respondents were asked whether they had smoked more than 100 cigarettes in their lifetimes. This variable is therefore binary, with '1' representing respondents that meet the criteria and '0' otherwise.

It is well known that excessive smoking has negative health impacts on individuals. A cross-sectional study conducted by @SmokingAndPhysicalActiveness found that smokers are less likely to be physically active compared to non-smokers. Hence, it would be interesting to discover if there is an effect of smoking more than 100 cigarettes on the time spent on physical activities.

@fig-barplot-cigarettes shows us that within the sample, the visual difference between the number of respondents who smoked more than 100 cigarettes (blue column) and respondents who smoked less than 100 cigarettes is not that big. Specifically, 57% of respondents have smoked less than 100 cigarettes while the remaining 43% of respondents smoked more than 100 cigarettes. As the classes in this variable are almost split in half.

@@

Initially, we fitted a full model containing every variable of interest. Since health region is a categorical variable with no implied order, To handle health region, we will be creating dummy variables for each health region, specifically $I*(HealthRegion_k=i)$ for $i \in {health\_region}$, and the reference chosen for health region would be the health_region = 35926 (The District of Algoma HU). We define $Income_k$ to be the personal income of individual k in our dataset.

@@

Overall, the individual t-test for our predictors was found to be below our set threshold of 95% alpha. The confidence interval for each predictor is listed in @tbl-confidence_interval_lmmodel shown below.
